---
title: "Concepts and Code Learned"
author: "Victoria Turkington"
date: "9/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 4 Course 1 R Markdown & Data Science Intro
### R Markdown

- bullets   [two spaces]
- **bold** 
- *italics* 
- ~~strikethrough~~ 
- [github.com](link.com)

### Types of Data Science Questions

- descriptive analysis: simple summaries of sample [US Census]
- exploratory analysis: explore relationships & help inform hypothesis generation
- inferential analysis: use small example to extrapolate to larger pop - requires thorough sampling scheme
- predictive analysis: using current and historical data to predict future outcomes [high data volume and simple modeling tends to do best here] predictive does not equal causation either
- causal analysis: manipulate independent variable with other factors controlled to determine cause [challenging to gather appropriate data]
- mechanistic analysis: exact changes in variables -> exact changes in other variables [physical or engineering science]

### Experimental Design

variables, confounders, p-hacking, control vs experimental groups, value of replication, sharing data, etc

### Big Data

volume, velocity, variety, 

structured data sets : typical excel sheet
unstructured: ex - emails, time spent on certain sites, etc

# Course 2
## Week 1 Writing Code
### Setting up Wd

```{r}
getwd()
```

- establish working directory - move as needed via misc/ change wd 
- to save environments/ functions open R ~~studio~~ text editor (button next to printer) 
- name them and save file

ex: change wd - coursera
```{r}
# dir() # lists files available
# source('test funct.R') # makes them accesible
# ls() # to see environments/ functions
# test.function(2) # should = 6
```

### R Console Input & Evaluation

```{r}
yn<- c(1, 1.5, 2, 3) # <- = assignment vector
yn # autoprint 
print(yn) # explicit print [1] means that is the first value in that vector row
msg<- "helloWorld"
msg
```

### R Data & Objects + Vectors & Lists

All objects can be classified as:
- numeric (real numbers)
- character 
- integer (type L after number)
- complex 
- logical (T/F) 

All vectors can only contain one type of the above objects
Exception: Lists are great for combining objects from different classes

All objects have various attributes
- names, dimnames  
- dimensions 
- classification 
- length for example if a vector has 8 values its length is 8 
- additional attributes as defined 

- Inf = Infinity  
- NaN: undefined object; can be used for missing values  
- vector("insert_class", length = insert ) # empty vector  
- ls() # list  
- attributes() # displays attributes  
- c() # concatenate values -> vector  

```{r}
yc<- c("a", "b", "c")
yc
attributes(yc)
yi<- vector("integer", length = 5)
yi
yl<- c(T, F)
ycomplex<- c(1+ 0i, 2+ 4i)
ycomplex
```

If you mix different classes together implicit coercion occurs;character & numeric -> characters, etc

Can explicity change class via as.insert_class(). Returns NA & Error if no logical interpretation

```{r}
as.numeric(yc)
as.integer(yl)
ylist<- list("a", 1, 4+2i, FALSE)
ylist ## dbl brackets signifies list! 
```

### Matrices

same class of objects required

```{r}
m<- matrix(nrow = 2, ncol = 4) ##empty matrix
m
m<- matrix(1:8, nrow = 2, 4) ## adds columnwise
m
dim(m) ## matrix dimensions
m<- 1:10
dim(m) <- c(2,5) ## transforms vector to matrix
m
m1<- c(1:5)
m2<- c(11:15)
m<- rbind(m1, m2) ##binds two vectors into matrix rows
m
m<- cbind(m1,m2) ##binds two vectors into matrix columns
m
```

### Factors!

used for categorical data to prescribe data labels/ levels; M/F, high, med, low, etc

```{r}
f<- factor(c("HS", "College", "Grad", "Grad", "HS", "HS"))
f
attributes(f) ## levels assigned alphabetically
f<- factor(c("HS", "College", "Grad", "Grad", "HS", "HS"),
           levels = c("HS", "College", "Grad")) ## manually determines levels
f
```
### Missing Values

NaNs are NA but NAs aren't NaNs

```{r}
n<- c(1, 2, NaN, NA, 5)
is.na(n)
is.nan(n)
```

### Data Frames

List in which all vectors have the same length! Can have an assortment of class types.
Special Attributes: row.names (can be subject names or just 1, 2, 3,)

Can be created by calling read.table() or read.csv() OR converted to matrix via data.matrix()

```{r}
B2M<- data.frame(child = c("boys", "boys"), number = c(2, 2), adult = c("men", "men"))
B2M
```

- 