---
title: "Concepts and Code Learned"
author: "Victoria Turkington"
date: "9/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 4 R Markdown & Data Science Intro
### R Markdown

- bullets   [two spaces]
- **bold** 
- *italics* 
- ~~strikethrough~~ 
- [github.com](link.com)

### Types of Data Science Questions

- descriptive analysis: simple summaries of sample [US Census]
- exploratory analysis: explore relationships & help inform hypothesis generation
- inferential analysis: use small example to extrapolate to larger pop - requires thorough sampling scheme
- predictive analysis: using current and historical data to predict future outcomes [high data volume and simple modeling tends to do best here] predictive does not equal causation either
- causal analysis: manipulate independent variable with other factors controlled to determine cause [challenging to gather appropriate data]
- mechanistic analysis: exact changes in variables -> exact changes in other variables [physical or engineering science]

### Experimental Design

variables, confounders, p-hacking, control vs experimental groups, value of replication, sharing data, etc

### Big Data

volume, velocity, variety, 

structured data sets : typical excel sheet
unstructured: ex - emails, time spent on certain sites, etc

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
