---
title: "Concepts and Code Learned"
author: "Victoria Turkington"
date: "9/29/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Course 1
## Week 4 Course 1 R Markdown & Data Science Intro
### R Markdown

- bullets   [two spaces]
- **bold** 
- *italics* 
- ~~strikethrough~~ 
- [github.com](link.com)

### Types of Data Science Questions

- descriptive analysis: simple summaries of sample [US Census]
- exploratory analysis: explore relationships & help inform hypothesis generation
- inferential analysis: use small example to extrapolate to larger pop - requires thorough sampling scheme
- predictive analysis: using current and historical data to predict future outcomes [high data volume and simple modeling tends to do best here] predictive does not equal causation either
- causal analysis: manipulate independent variable with other factors controlled to determine cause [challenging to gather appropriate data]
- mechanistic analysis: exact changes in variables -> exact changes in other variables [physical or engineering science]

### Experimental Design

variables, confounders, p-hacking, control vs experimental groups, value of replication, sharing data, etc

### Big Data

volume, velocity, variety, 

structured data sets : typical excel sheet
unstructured: ex - emails, time spent on certain sites, etc

# Course 2
## Week 1 Writing Code
### Setting up Wd

```{r}
getwd()
```

- establish working directory - move as needed via misc/ change wd 
- to save environments/ functions open R ~~studio~~ text editor (button next to printer) 
- name them and save file

ex: change wd - coursera
```{r}
# dir() # lists files available
# source('test funct.R') # makes them accesible
# ls() # to see environments/ functions
# test.function(2) # should = 6
```

### R Console Input & Evaluation

```{r}
yn<- c(1, 1.5, 2, 3) # <- = assignment vector
yn # autoprint 
print(yn) # explicit print [1] means that is the first value in that vector row
msg<- "helloWorld"
msg
```

### R Data & Objects + Vectors & Lists

All objects can be classified as:
- numeric (real numbers)
- character 
- integer (type L after number)
- complex 
- logical (T/F) 

All vectors can only contain one type of the above objects
Exception: Lists are great for combining objects from different classes

All objects have various attributes
- names, dimnames  
- dimensions 
- classification 
- length for example if a vector has 8 values its length is 8 
- additional attributes as defined 

- Inf = Infinity  
- NaN: undefined object; can be used for missing values  
- vector("insert_class", length = insert ) # empty vector  
- ls() # list  
- attributes() # displays attributes  
- c() # concatenate values -> vector  

```{r}
yc<- c("a", "b", "c")
yc
attributes(yc)
yi<- vector("integer", length = 5)
yi
yl<- c(T, F)
ycomplex<- c(1+ 0i, 2+ 4i)
ycomplex
```

If you mix different classes together implicit coercion occurs;character & numeric -> characters, etc

Can explicity change class via as.insert_class(). Returns NA & Error if no logical interpretation

```{r}
as.numeric(yc)
as.integer(yl)
ylist<- list("a", 1, 4+2i, FALSE)
ylist ## dbl brackets signifies list! 
```

### Matrices

same class of objects required

```{r}
m<- matrix(nrow = 2, ncol = 4) ##empty matrix
m
m<- matrix(1:8, nrow = 2, 4) ## adds columnwise
m
dim(m) ## matrix dimensions
m<- 1:10
dim(m) <- c(2,5) ## transforms vector to matrix
m
m1<- c(1:5)
m2<- c(11:15)
m<- rbind(m1, m2) ##binds two vectors into matrix rows
m
m<- cbind(m1,m2) ##binds two vectors into matrix columns
m
```

### Factors!

used for categorical data to prescribe data labels/ levels; M/F, high, med, low, etc

```{r}
f<- factor(c("HS", "College", "Grad", "Grad", "HS", "HS"))
f
attributes(f) ## levels assigned alphabetically
f<- factor(c("HS", "College", "Grad", "Grad", "HS", "HS"),
           levels = c("HS", "College", "Grad")) ## manually determines levels
f
```
### Missing Values

NaNs are NA but NAs aren't NaNs

```{r}
n<- c(1, 2, NaN, NA, 5)
is.na(n)
is.nan(n)
```

### Data Frames

List in which all vectors have the same length! Can have an assortment of class types.
Special Attributes: row.names (can be subject names or just 1, 2, 3,)

Can be created by calling read.table() or read.csv() OR converted to matrix via data.matrix()

```{r}
B2M<- data.frame(child = c("boys", "boys"), number = c(2, 2), adult = c("men", "men"))
B2M
nrow(B2M)
ncol(B2M)
```

### Names

useful for creating self describing data!

Vector Examples

```{r}
v<- 1:3
names(v)<- c("a", "b", "c")
v
```

List Examples

```{r}
l<- list("a"=1, "b"=2, "c"=3)
l
```

Matrix Examples

```{r}
m<- matrix(1:4, nrow = 2, ncol = 2)
m
dimnames(m)<- list(c("c", "d"), c("a", "b")) #row, column
m
```

### Reading Tabular Data

Reading | Writing
read.csv/ read.table | write.table
readLines | writeLines  (text as character vector)
source | dump (R code files - functions, etc)
dget |dput (R code files deparsed into text files)
load | save (reading in saved workspaces)
unserialize | serialize (reading single R objects in binary form)

read.table(file, header, sep, colClasses, nrows, comment.char, skip, stringsAsFactors)
*see help page for more info

Lots of info about outside connections - view as needed.

### Subsetting

- Basics  
- Lists 
- Matrices 
- Partial Matching 
- Removing missing values 

**Basics**
- [] returns object of the same class as the original (vector -> vector); can select multiple elements of an object  
- [[]] extract ONE ELEMENT from list/ data frame -> returned object may be of different class  
- $ extract element of object by name otherwise similar to [[]]

```{r}
b<- c(1, 4, 7, 9, 9, 4, 5)
b
b[1]
b[3]
b[1:5]
b[b>5]
u<- b < 7
u
```

**Lists**

```{r}
l<- list(me = "hungry", cookies = 1:8, chubbybunny = "you")
l[2] ## gives list
l[[2]] ## gives sequence
l$cookies
l[1]
l[[1]]
l$me
l[c(1,3)] ## single bracket for multiple elems
```

computed indicies; [[]]

```{r}
f<- function(x) x^2
l<- list(atticus = f(2:5), boo = "radley") 
finch<- "atticus"
l[[finch]] ## returns computed index
l[finch] 
l$finch ## null b/c looking for name w/in list == "finch"
l<- list(go = list(5,6,7,8), team = list("who", "do", "we", "appreciate"))
l[[c(1,3)]]
l[[c(2,4)]]
```

**Matrices**

```{r}
m<- matrix(1:10, nrow = 5, ncol = 2)
m[4,1] ## Returns vector
m[4,1,drop = FALSE] ## Returns matrix (prevents 'drop' of dimensions)
m[4,]
m[4, ,drop = FALSE] 
```

**Partial Matching**
```{r}
l$te ## does not work for [[]]
```

**Removing Missing Values **
```{r}
n<- c("a", "b", NA, "d")
na<- is.na(n)
n[na] ##get whats missing
n[!na] ##opposite of what's missing
n1<- c(1, 2, 3, NA)
good<- complete.cases(n, n1)
n[good] #which elements between two are not missing any data? -> only 1&2
n1[good]
```

**Vectorized Operations**

actions occur respectively
```{r}
x<- c(1:5)
y<- c(6:10)
x*y
x>=3
m<- matrix(1:4, nrow =2, ncol=2)
m1<- matrix(5,4, nrow=2, ncol=2) ## produces matrix w/ "5" values in all four quadrants
m*m1
m/m1
m%*%m1 ## true matrix multiplication
```

```{r}
x<- 1:4
x[x %in% 1:2] <- 0
x
```

### Random Wk 1 Helpful Functions
```{r}
args(list.files)
seq(1, 10, by = .5)
x<- seq(from= 4, to = 7, length.out = 12)
x
x[-c(2, 10)]
y<- rep(c(5,6,7,8), times = 3)
y
rep(c(5,6,7,8), each = 3)
paste(x, y, collapse = "-")
```

## Week 2 - Control Structures & Functions

Control functions are useful for building functions - not necessarily for typing in the command line

### Control Structures

#### If/Else

If: perform an action given a specified condition is true
else: perform alternative action if condition is not true
Command can contain multiple conditions
Command does not require "else"

```{r}
x<- 3+0
if(0<x & x>=3) {y<- "Low"}
if(3<x & x<=7) {y<- "Medium"}
if(7<x & x<=10) {y<- "High"}
```

#### For
Loop index == i or j,k,l,etc

```{r}
for(i in 1:10) {
  print(i)}
x<- c("a", "b", "c")
for(letter in x) {
  print(letter)
} ## can omit curly braces if single expression in body

## Be careful w/ nested for loops for matrices (more than 2-3 levels can appear quite confusing)
```

### While

can be useful and simple commands
**beware** of infinite looping - need to make sure convergence occurs at some point!
generally safer to use for command because it comes with a hard limit of runs!

```{r}
count<- 0
while(count <10) {
  print(count)
  count<- count + 1
  }
```

### Repeat, Next, & Retutn

#### Repeat:
initiates infinite loop : only way to end a repeat function is with "break"
relatively limited statistical application
generally used to optimization problems

#### Next:
Skips output for given condition : in this case - skips 1-5
```{r}
for (i in 1:10) {
  if(i <= 5) {
    next
    }
  print(i)
}
```

#### Return
signals that a function should exit & return given value
```{r}
for (i in 1:10) {
  if(i == 5) {
    return("5")
  }
  print(i)
}
```

### Coding Standards

- always save code to a text file  
- ident (limit code width to 80 columns)
  - rstudio/preferences/code/ tab width :: command A(select all)/ command I (applies indent)
  - 8 space indents
- limit length of functions 
  - be intentional about what you assign to what function 
  - helpful for finding errors 

### Dates & Times

Dates are represented by "Date" class (days since 1/1/1970)

```{r}
x<- as.Date("1970-01-01")
x
x+1
unclass(x)
unclass(as.Date("1971-01-01"))
```

Times are rep by POSIXct or POSIXlt class (seconds since 1/1/1970)
- ct: very large integer (helpful when wanting to store times in df)
- lt: list which also stores relevant date info

Relevant generic funcs for DATE & TIME : weekdays, months, quarters

```{r}
x<- Sys.time()
x
p<- as.POSIXlt(x)
names(unclass(p))
p$sec
y<- Sys.time() ## automatically in POSIXct
unclass(y) ## produces integer accordingly
```

strptime() is a great function for creating data frames with time/ date info! just check help page for use its very simple.
can do mathematical operations with vectors but they need to be of the same class!
even keeps track of leap years, daylight savings, time zone (tz = ), etc.

## Week 3 Loop Functions & Debugging

###Loop functions
- lapply: loops over list and applies function to each elem  
- sapply: same as ^ but simplified output - if elems have length 1 -> vector; if all same length -> matrix
- apply: function to margins of an array (helpful for data w/ dimensions)
- tapply: function to subsets of a vector (table apply) 
- mapply: multivariate version of lapply 
- split: also helpful esp. w/ lapply 

```{r}
l<- list(a = 1:3, b = rnorm(10))
lapply(l, mean) ## (vector aka list, function)
lapply(l, runif) ## runif -> random number generator for uniform distrib given designated number of values
lapply(l, runif, min = 0, max = 10) ## runif expanded from 0-1 to 0-10

ml<- list(a = matrix(1:6, nrow = 2, ncol = 3), b = matrix(1:8, nrow = 2, ncol = 4))
lapply(ml, function(anonymous) anonymous[1,]) ## created an anon function -> extracts first row of each
```
 Apply
 
```{r}
y<- matrix(rnorm(100), nrow = 5, ncol = 20)
apply(y, 2, mean) ## Provides mean of each column ; collapses 1st dimension and KEEPS the SECOND!!
apply(y, 1, mean) ## Provides mean of each row ; collapses 2nd dimension and KEEPS the FIRST!!
apply(y, 1, quantile, probs = c(0.25, 0.75))
a<- array(rnorm(2*2*10), c(2,2,10))
apply(a, c(1,2), mean) ## collapses third dimension of array
rowMeans(a, dims = 2)
```

shortcuts: (faster execution)
rowMeans
rowSums
colMeans
colSums

Mapply
applies function in parallel over argument set
can easily vectorize a function which doesn't typically take vector arguments
```{r}
list(rep(1,4), rep(2, 3), rep(3, 2), rep(4, 1)) ## We can see which code is more efficient quite clearly
mapply(rep, 1:4, 4:1)
```

Tapply

used to apply functions to subsets of a vector

tapply(x, index, function)
x : vector
index : factor --> subset

```{r}
x<- c(rnorm(10), runif(10), rnorm(10, 1))
f<- gl(3, 10)
t<- tapply(x, f, mean)
t ## autosimplifies -> can specify false (simplify = FALSE) to produce a list
tapply(x, f, range) ## list produced b/c cannot easily simplify

```

Split as a useful tool!
```{r}
split(x, f)
lapply(split(x,f), mean)
x<- rnorm(10)
f1<- gl(2, 5)
f2<- gl(5, 2)
interaction(f1,f2) ## can create empty levels (these are all theoretical possibilities)
str(split(x, list(f1, f2)))
str(split(x, list(f1, f2), drop = TRUE)) ## drops empty levels
```

### Debugging!
#### Conditions
- message: lets you know something unexpected occurred  
- warning: lets you know something when wrong but it is not fatal  
- error: fatal problem occurred  
**can create additional conditions!**

**Debugging tools**
- traceback : prints function call stack ; where in seq of function calls the error occurred
- debug : give function as argument step through func line by line
- browser : execution of specific function suspended rather than line by line
- trace : allows insertion of debugging code into function (can insert into someone elses function)
- recover : error handler function (stops execution of func immediately where error occurs)

## Week 4
### Str Function!
compactly displays internal structure of an R object (lovely for data frames)
- can be used in place of summary  
- great for compactly displaying large lists (incl nested lists)  
-  ~ 1 line per basic object  

```{r}
x<- rnorm(100, 2, 4)
summary(x)
str(x) ## gives number of values and first five values
f<- gl(40,10)
str(f) ## gives description and first several values
summary(f)
```
 
### Simulation

Generating Random numbers
- _norm : ___ w/ normal distribution
  - rnorm: generates variates w/ specified mean & SD
  - dnorm:  evaluates norm prob density at a point / vector of points
   - pnorm: eval cumulative distrib function for norm distrib
   - qnorm: quantile function
- rpois: random poisson variates w/ given rate

```{r}
set.seed(2) ## pick integer ; fixes same seq of "random" variables
rnorm(6)
rnorm(8)
set.seed(2)
rnorm(8) ## see that first six indicate same sequencing

rpois(5, 10) ## 5 integers; "rate" =10 ~ similar to mean
ppois(2, 3) ## prob x (aka 3)<=2 
ppois(7, 3) ## prob 3 <= 7
```

Sampling

set seed for reproducibility!!

```{r}
sample(1:10, 5) ## pick five random values from data 1:10
sample(1:10, 10, replace = TRUE) ## picks values w/ replacement
sample(letters, 3)

```

### R Profiling

optimizing code to perform more quickly if very time consuming
create b4 optimizing (otherwise you'll introduce bugs/ may run into issues)

system.time() : produces the amount of time it takes (s) to execute function/ evaluate an expression
  - for long expressions : check parts using curly braces 
-> returns object of class proc_time
  - user time (CPU charged time)
  - elapsed time (time experienced relative to actual time)
  
  if elapsed > user : if CPU spends lots of time waiting around for external issues
  if elapsed < user : program exploits additional cores
  - parallel processing : via parallel pkg
  
  Challenge: expression assumes you know where the problem is
  
Rprof : begins profiler
summaryRprof() : makes it readable

DO NOT USE w/ system.time() : it will bum you out

by.total(): spent in top level function (not where real work occurs)
by.self(): spent in given function (after subtracting out lower level functions)
